---
layout: post
title: Preparing some genomic annotations
tags: genome
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)
library(GenomicRanges)
library(data.table)
library(PopSV)
library(knitr)
```

## Mappability track

A mappability track was produced from the UCSC track. The raw file contains, for each base in the genome, an estimation of how likely a read is correctly mapped to this position. 

Using a sliding-window approach, I compute the average mappability in regions of size 1 Kbp.
This is a more manageable amount of data and still informative, especially when interested in large regions (e.g. SVs).

I used a custom Perl script to efficiently parse the bedGraph-transformed original file. See the code on [GitHub](https://github.com/jmonlong/Hippocamplus/blob/gh-pages/_source/mappabilityBin.pl).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
download.file("http://hgdownload.cse.ucsc.edu/admin/exe/linux.x86_64.v287/bigWigToBedGraph","bigWigToBedGraph")
system2("chmod","+x bigWigToBedGraph")
download.file("http://hgdownload.soe.ucsc.edu/gbdb/hg19/bbi/wgEncodeCrgMapabilityAlign100mer.bw","wgEncodeCrgMapabilityAlign100mer.bw")
system2("./bigWigToBedGraph","wgEncodeCrgMapabilityAlign100mer.bw wgEncodeCrgMapabilityAlign100mer.bed")
system2("perl","mappabilityBin.pl -i wgEncodeCrgMapabilityAlign100mer.bed -o map100mer-1kbp.bed -bin 1000")
system2("gzip","-f map100mer-1kbp.bed")
rm = file.remove(c("bigWigToBedGraph", "wgEncodeCrgMapabilityAlign100mer.bw", "wgEncodeCrgMapabilityAlign100mer.bed"))
```

I uploaded the result there: [https://dl.dropboxusercontent.com/s/i537zjs65dpw34n/map100mer-1kbp.bed.gz?dl=0](https://dl.dropboxusercontent.com/s/i537zjs65dpw34n/map100mer-1kbp.bed.gz?dl=0).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
map.df = as.data.frame(fread("gunzip -c map100mer-1kbp.bed.gz", showProgress = FALSE))
colnames(map.df) = c("chr", "start", "end", "map")
map.df %<>% mutate(chr=factor(gsub("chr","",chr), levels=c(1:22,"X","Y"))) %>% filter(!is.na(chr))
```

We can cut the genome into three mappability classes:

+ **unique** regions with high mappability estimate (>0.95).
+ **low-map** regions with a non-null mappability but lower than 0.95.
+ **no-map** regions with mappability 0.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
map.df %<>% mutate(map.class=ifelse(map>.95,"unique", "low-map"), map.class=ifelse(map==0,"no-map", map.class))
ggplot(map.df, aes(x=map, fill=reorder(map.class, map.class, length))) + geom_histogram() + theme_bw() + ylab("number of 1 kb bins") + scale_fill_brewer(palette="Set1", name="map class")
map.s = map.df %>% group_by(map.class) %>% summarize(Mb=n()/1000) %>% ungroup %>% mutate(prop=Mb/sum(Mb)) %>% arrange(desc(Mb))
map.s %>% kable(digits=3)
```
