---
title: Word Cloud in R
date: 2016-02-26
tags: ["R", "plot"]
---

```{r, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=50))
```

The [`wordcloud` package](https://cran.r-project.org/web/packages/wordcloud/index.html) is available on CRAN.

## Fake words

I create fake words to see a bit how the command is working.

```{r example}
library(wordcloud)
createWords <- function(w.l=3) paste(sample(letters,w.l,TRUE),collapse="")
words = sapply(1:200,function(e)createWords(runif(1,3,10)))
freq = c(sample(1:30,190,T),sample(30:150,10,T))
freq = freq / sum(freq)
wordcloud(words,freq)
## Big words in the center
wordcloud(words,freq, random.order=FALSE)
## Max word number
wordcloud(words,freq, max.words = 50)
## Rotation: proportion of 90 degree
wordcloud(words,freq, rot.per = .01)
## Color the most frequent words
wordcloud(words, freq, colors=c('black','blue','red'), random.order=FALSE)
## Color for each word
wordcloud(words, freq, colors=sample(c('black','blue','red'), length(words), TRUE), random.order=FALSE, ordered.colors = TRUE)
```

## Command history

I retrieved the commands from my `.bash_history` files (laptop and HPCs) and I want to make a word cloud showing the commands I use the most.

```{r cmds}
library(dplyr)
library(RColorBrewer)
cmds = read.table('../../data/bash-commands.tsv.gz', as.is=TRUE)
colnames(cmds) = c('cmd', 'machine')
```

### Laptop

By default the maximum history size was set to 500 commands so I don't have the full set of commands, just the last 500. (I increased the limit, see you in 10,000 commands.)

```{r cmdlt}
cmds.s = cmds %>% filter(machine=='laptop', !grepl('=',cmd), !grepl('\\.',cmd), !grepl('/',cmd)) %>% group_by(cmd) %>% summarize(n=n()) %>% mutate(freq=n/sum(n))
wordcloud(cmds.s$cmd, cmds.s$freq, colors=c('black', brewer.pal(8, 'Set1')), random.order=FALSE, scale=c(10,1), title='All')
```

### HPC

```{r cmdhpc}
cmds.s = cmds %>% filter(machine!='laptop', !grepl('=',cmd), !grepl('\\.',cmd), !grepl('/',cmd), !grepl('\\$',cmd), !grepl('\\:',cmd)) %>% group_by(cmd) %>% summarize(n=n()) %>% mutate(freq=n/sum(n))
wordcloud(cmds.s$cmd, cmds.s$freq, colors=c('black', brewer.pal(8, 'Set1')), random.order=FALSE, scale=c(10,1), title='All', min.freq=1)
```

Good see the usual suspects `ls` and `cd` and their "typo" versions `;s`/`ks`/`ld` and `xs`.
