---
title: Segmental duplication exploration
date: 2016-10-20
tags: ["genome"]
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=8)
```

```{r libs}
library(ggplot2)
library(dplyr)
library(magrittr)
library(knitr)
library(GenomicRanges)
library(rtracklayer)
## library(waffle)
library(RColorBrewer)
## library(ggrepel)
library(PopSV)
library(clusterProfiler)
## library(data.table)
winsor <- function(x,u=10){
    if(any(x>=u)) x[x>=u] = u
    x
}
olBins <- function(bins.df, feat.df, label="ol"){
  bins.gr = makeGRangesFromDataFrame(bins.df)
  feat.gr = makeGRangesFromDataFrame(feat.df)
  res.df = data.frame(n=rep(0,nrow(bins.df)), d=rep(0,nrow(bins.df)))
  colnames(res.df) = paste(label,colnames(res.df), sep=".")
  res.df[,1] = countOverlaps(bins.gr, feat.gr)
  feat.gr = reduce(feat.gr)
  ol = findOverlaps(bins.gr, feat.gr) %>% as.data.frame %>% mutate(qsw=width(pintersect(bins.gr[queryHits], feat.gr[subjectHits]))) %>% group_by(queryHits) %>% summarize(bp=sum(qsw))
  res.df[ol$queryHits,2] = ol$bp
  res.df[,2] = res.df[,2] / width(bins.gr)
  bins.df = cbind(bins.df, res.df)
  bins.df
}
## library(BSgenome.Hsapiens.UCSC.hg19)
## seql.1.22 = seqlengths(BSgenome.Hsapiens.UCSC.hg19)[paste0("chr",1:22)]
## rand.gr <- function(widths){
##     chrs = apply(rmultinom(length(widths), 1, (seql.1.22/1e3)/sum(seql.1.22/1e3)),2,function(e)which(e==1))
##     starts = runif(length(widths), 0, seql.1.22[chrs]-widths)
##     return(GRanges(chrs, IRanges(starts, width=widths)))
## }
```

## Segmental Duplications (SD)

```{r sddl}
if(!file.exists("genomicSuperDups.txt.gz")) download.file("http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/genomicSuperDups.txt.gz", "genomicSuperDups.txt.gz")
sd = read.table('genomicSuperDups.txt.gz', as.is=TRUE)
sd = sd[,c(2:4,8:10,27)]
colnames(sd) = c('chr', 'start', 'end', 'chr2', 'start2', 'end2', 'fracMatch')
sd %<>% mutate(size=end-start)
```

I downloaded the segmental duplication annotation for hg19 from [UCSC](http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/genomicSuperDups.txt.gz). There are `r nrow(sd)` annotated SD.
They are defined as regions larger than 1 Kbp with at least 90% similarity with another region in the genome.

## Segmental duplication regions

Many SD are nested of located next to each other. 
I merge overlapping SDs (or located at <10 bp) to create *SD regions*, i.e. longer stretch of the genome overlapping SDs.

```{r sdreg}
sdreg = makeGRangesFromDataFrame(sd) %>% reduce(min.gapwidth=10)
```

There are `r length(sdreg)` SD regions, that account for `r round(sum(width(sdreg)/1e6),1)` Mbp of the genome.


## Size distribution


```{r size}
ggplot(sd, aes(size)) + geom_histogram() + theme_bw() + scale_x_log10(breaks=c(1e3,1e4,1e5,1e6)) + xlab('size (bp)') + ylab('segmental duplication')
sdreg %>% as.data.frame %>% ggplot(aes(width)) + geom_histogram() + theme_bw() + scale_x_log10(breaks=c(1e3,1e4,1e5,1e6)) + xlab('size (bp)') + ylab('segmental duplication region')
```

## Similarity

```{r sim}
sd %<>% mutate(sim.class=cut(fracMatch, seq(.9, 1, .02), include.lowest = TRUE))
ggplot(sd, aes(x=fracMatch, fill=sim.class)) + geom_histogram() + theme_bw() + xlab('similarity') + ylab('segmental duplication') + scale_fill_brewer(palette='Set2', name='similarity class') + theme(legend.position=c(.99,.99), legend.justification=c(1,1))
```

## Chromosome distribution

```{r chr}
gfrag.df = fragment.genome.hg19(1e5, XY.chr = TRUE)
gfrag.df %<>% mutate(chr=factor(chr, levels=c(1:22,"X","Y")))
sd.d = sd %>% mutate(chr=gsub('chr','',chr)) %>% group_by(sim.class) %>% do(olBins(gfrag.df, ., "sd"))
sd.d %>% group_by(chr, sim.class) %>% summarize(sd.d=mean(sd.d)) %>% ggplot(aes(x=chr, y=sd.d, colour=sim.class, group=sim.class)) + geom_point() + geom_line() + theme_bw() + ylab("average density of SD") + xlab('chromosome') + theme(legend.position=c(.01,.99), legend.justification=c(0,1)) + scale_colour_brewer(palette='Set2', name='similarity class') 
sd.d %>% mutate(chr=factor(chr, levels=rev(levels(chr)))) %>% ggplot(aes(x=start, y=chr)) + geom_tile(aes(fill=sd.d)) + theme_bw() + xlab("position") + scale_fill_gradient(name="SD\ndensity") + theme(legend.position=c(.99,.01), legend.justification=c(1,0))
```

A few chromosomes are more enriched in SD. Some have long stretches of SD, e.g. chr 9 or chr Y. These peaks are mostly created with very recent/similar SDs.


## Distance to the other segment

```{r dist}
sd %>% mutate(otherSeg=ifelse(chr==chr2, 'same chr', 'different chr')) %>% group_by(otherSeg) %>% summarize(segdup=n()) %>% ungroup %>% mutate(segdup.prop=segdup/sum(segdup)) %>% kable(digits=3)
sd %>% filter(chr==chr2) %>% mutate(d1=start2-end, d2=end2-start, d=ifelse(d1>d2,d2,d1)) %>% ggplot(aes(x=d)) + geom_histogram() + theme_bw() + scale_x_log10() + xlab('distance to similar region (bp)') + ggtitle('SD with similar region in same chromosome')
```

For the majority of SDs, the similar region is in a different chromosome.
For the others the majority are far from each other.


## Gene content

```{r gcdl, cache=TRUE}
if(!file.exists("gencode.v19.gtf.gz")) download.file("ftp://ftp.sanger.ac.uk/pub/gencode/Gencode_human/release_19/gencode.v19.annotation.gtf.gz", "gencode.v19.gtf.gz")
gtf = import("gencode.v19.gtf.gz")
gtf = gtf[as.character(seqnames(gtf)) %in% paste0('chr', c(1:22,'X','Y'))]
genes = subset(gtf, type=='gene')
```

I downloaded Gencode v19 at [ftp://ftp.sanger.ac.uk/pub/gencode/Gencode_human/release_19/gencode.v19.annotation.gtf.gz](ftp://ftp.sanger.ac.uk/pub/gencode/Gencode_human/release_19/gencode.v19.annotation.gtf.gz).


Around a thousand protein-coding genes are completely within SD regions. 

```{r gene}
gene.sd = genes[overlapsAny(genes, sdreg, type='within')]
gene.sd %>% as.data.frame %>% select(gene_type, gene_name) %>% unique %>% group_by(gene_type) %>% summarize(gene=n()) %>% arrange(desc(gene)) %>% kable
```

### Gene families

A naive way of looking for gene families is to cluster the gene names. 
I also remove any trailing numbers in the gene name.

```{r genefam}
genen = subset(gene.sd, gene_type=='protein_coding')$gene_name %>% unique
genen = gsub('[0-9]*$', '', genen)
gened = adist(genen)
colnames(gened) = rownames(gened) = genen
hc = hclust(as.dist(gened), method='ward.D')
genef = cutree(hc, h=2)
data.frame(gene=genen, fam=genef) %>% group_by(fam) %>% summarize(example=gene[1], n=n()) %>% arrange(desc(n), fam) %>% select(-fam) %>% filter(n>9) %>% kable
```

### Gene Ontology

```{r go, cache=TRUE}
genes.all = unique(subset(genes, gene_type=='protein_coding')$gene_name)
eg = bitr(genes.all, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
eg$sd = eg$SYMBOL %in% subset(gene.sd, gene_type=='protein_coding')$gene_name
go.enr.bp = enrichGO(gene=subset(eg, sd)$ENTREZID, 'org.Hs.eg.db', ont="BP", universe=eg$ENTREZID, readable=TRUE)
go.enr.cc = enrichGO(gene=subset(eg, sd)$ENTREZID, 'org.Hs.eg.db', ont="CC", universe=eg$ENTREZID, readable=TRUE)
go.enr.mf = enrichGO(gene=subset(eg, sd)$ENTREZID, 'org.Hs.eg.db', ont="MF", universe=eg$ENTREZID, readable=TRUE)
```


```{r gographs, fig.height=10, fig.width=20}
## dotplot(go.enr, showCategory=20)
go.enr.bp.s = summary(go.enr.bp)
go.enr.bp.s[,c("Description","GeneRatio","qvalue")] %>% kable

go.enr.mf.s = summary(go.enr.mf)
go.enr.mf.s[,c("Description","GeneRatio","qvalue")] %>% kable

go.enr.cc.s = summary(go.enr.cc)
go.enr.cc.s[,c("Description","GeneRatio","qvalue")] %>% kable

## enrichMap(go.enr, vertex.label.cex=1)
```
