---
title: Repeat Masker exploration
date: 2016-06-04
tags: ["genome"]
draft: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=8)
```

```{r libs}
library(ggplot2)
library(dplyr)
library(magrittr)
library(knitr)
library(GenomicRanges)
library(rtracklayer)
library(waffle)
library(RColorBrewer)
## library(ggrepel)
library(PopSV)
library(data.table)
winsor <- function(x,u=10){
    if(any(x>=u)) x[x>=u] = u
    x
}
olBins <- function(bins.df, feat.df, label="ol"){
  bins.gr = makeGRangesFromDataFrame(bins.df)
  feat.gr = makeGRangesFromDataFrame(feat.df)
  res.df = data.frame(n=rep(0,nrow(bins.df)), d=rep(0,nrow(bins.df)))
  colnames(res.df) = paste(label,colnames(res.df), sep=".")
  res.df[,1] = countOverlaps(bins.gr, feat.gr)
  feat.gr = reduce(feat.gr)
  ol = findOverlaps(bins.gr, feat.gr) %>% as.data.frame %>% mutate(qsw=width(pintersect(bins.gr[queryHits], feat.gr[subjectHits]))) %>% group_by(queryHits) %>% summarize(bp=sum(qsw))
  res.df[ol$queryHits,2] = ol$bp
  res.df[,2] = res.df[,2] / width(bins.gr)
  bins.df = cbind(bins.df, res.df)
  bins.df
}
## library(BSgenome.Hsapiens.UCSC.hg19)
## seql.1.22 = seqlengths(BSgenome.Hsapiens.UCSC.hg19)[paste0("chr",1:22)]
## rand.gr <- function(widths){
##     chrs = apply(rmultinom(length(widths), 1, (seql.1.22/1e3)/sum(seql.1.22/1e3)),2,function(e)which(e==1))
##     starts = runif(length(widths), 0, seql.1.22[chrs]-widths)
##     return(GRanges(chrs, IRanges(starts, width=widths)))
## }
```

## Repeat Masker

```{r dl}
if(!file.exists("rmsk.txt.gz")) download.file("http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/rmsk.txt.gz","rmsk.txt.gz")
rm.df = as.data.frame(fread("gunzip -c rmsk.txt.gz", showProgress = FALSE, sel=c(6:8, 11:13)))
colnames(rm.df) = c("chr","start","end","repName","repClass","repFamily")
rm.df %<>% mutate(chr=gsub("chr","",chr)) %>% filter(chr %in% c(1:22, "X", "Y"))
rm.gr = makeGRangesFromDataFrame(rm.df)
if(!file.exists("genomicSuperDups.txt.gz")) download.file("http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/genomicSuperDups.txt.gz", "genomicSuperDups.txt.gz")
sd = read.table('genomicSuperDups.txt.gz', as.is=TRUE)
sd = sd[,c(2:4,8:10,27)]
colnames(sd) = c('chr', 'start', 'end', 'chr2', 'start2', 'end2', 'fracMatch')
sd %<>% mutate(chr=gsub("chr","",chr)) %>% filter(chr %in% c(1:22, "X", "Y"))
sd.gr = makeGRangesFromDataFrame(sd)
```

I downloaded Repeat Masker at [http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/rmsk.txt.gz](http://hgdownload.soe.ucsc.edu/goldenPath/hg19/database/rmsk.txt.gz).

```{r ex}
rm.df %>% sample_n(10) %>% kable(row.names=FALSE)
```

The annotation spans a total of `r round(sum(width(reduce(rm.gr))/1e9),2)` Gbp. 
If we include segmental duplication, `r round(sum(width(reduce(c(rm.gr, sd.gr)))/1e9),2)` Gbp are annotated as repeat in the human genome.


### TE Density

How many 500bp windows contain a TE ?

```{r tedens}
bins500 = fragment.genome.hg19(500)
tes = c("LINE","DNA","SINE","LTR","DNA?","Other","SINE?","LTR?","LINE?")
bins500$te = overlapsAny(makeGRangesFromDataFrame(bins500), makeGRangesFromDataFrame(subset(rm.df, repClass %in% tes)))
mean(bins500$te)

bins1000 = fragment.genome.hg19(1000)
bins1000$te = overlapsAny(makeGRangesFromDataFrame(bins1000), makeGRangesFromDataFrame(subset(rm.df, repClass %in% tes)))
mean(bins1000$te)
```

### Alpha satellites

```{r alpha}
alpha = subset(rm.df, repName=='ALR/Alpha') %>% makeGRangesFromDataFrame
mean(width(alpha))
```


### Number

Focusing on autosomes/X/Y, there are `r format(length(genes), big.mark=',')` "genes" of different types:

```{r genetypes}
pal = c(brewer.pal(9,'Set1'), brewer.pal(8,'Set2'))[-6]
genes %>% as.data.frame %>% mutate(gene_type=ifelse(grepl('IG|TR', gene_type), 'IG/TR', gene_type)) %>% group_by(gene_type) %>% summarize(vals=ceiling(n()/100)) %>% dplyr::rename(names=gene_type) %>% arrange(desc(vals)) %>% as.data.frame %>% waffle(colors=pal[1:15], legend_pos='bottom', rows=15, xlab='1 sq = 100 genes')
```

I merge the rare types into a *other* class and some RNAs.

```{r genetypestable}
genes$gene_type.f = "other"
genes$gene_type.f[grepl("RNA", genes$gene_type)] = "RNA"
genes$gene_type.f[genes$gene_type%in%c("pseudogene","lincRNA","protein_coding","miRNA")] = genes$gene_type[genes$gene_type%in%c("pseudogene","lincRNA","protein_coding","miRNA")]
genes %>% as.data.frame %>% group_by(gene_type.f) %>% summarize(n=n()) %>% arrange(desc(n)) %>% kable
```

### Size

```{r genesize}
genes %>% as.data.frame %>% ggplot(aes(x=width, fill=gene_type.f)) + geom_histogram() + ylab("number of genes") + xlab("gene length (bp)") + theme_bw() + scale_x_log10(breaks=10^(1:6)) + scale_fill_brewer(palette="Set1") + theme(legend.position=c(.99,.99), legend.justification=c(1,1), legend.title=element_blank())
```

The largest annotated genes span more than 2 Mbp:

```{r largegenes}
genes %>% as.data.frame %>% arrange(desc(width)) %>% head(5) %>% mutate(size.Mbp=(width)/1e6) %>% select(seqnames, gene_name, gene_type, size.Mbp) %>% kable
```

The smallest protein-coding annotated genes less than 100 bp:

```{r smallgenes}
genes %>% as.data.frame %>% filter(gene_type=="protein_coding") %>% arrange(width) %>% head(5) %>% mutate(size.bp=width) %>% select(seqnames, gene_name, gene_type, size.bp) %>% kable
```

### Density

Using non-overlapping windows of 1 Mb the gene density looks like this:

```{r genedens}
gfrag.df = fragment.genome.hg19(1e6, XY.chr = TRUE)
gfrag.df %<>% mutate(chr=factor(chr, levels=c(1:22,"X","Y")))
gene.d = genes %>% as.data.frame %>% mutate(seqnames=gsub('chr','',seqnames)) %>% group_by(gene_type.f) %>% do(olBins(gfrag.df, ., "gene"))
gene.d %>% group_by(chr, gene_type.f) %>% summarize(gene.n=mean(gene.n)) %>% ggplot(aes(x=chr, y=gene.n, colour=gene_type.f, group=gene_type.f)) + geom_point() + geom_line() + theme_bw() + ylab("average number of genes per Mb") + scale_colour_brewer(palette="Set1") + theme(legend.position=c(.01,.99), legend.justification=c(0,1), legend.title=element_blank()) + xlab('chromosome')
gene.d %>% filter(gene_type.f=="protein_coding") %>% mutate(chr=factor(chr, levels=rev(levels(chr)))) %>% ggplot(aes(x=start, y=chr)) + geom_tile(aes(fill=gene.n)) + theme_bw() + xlab("position") + geom_tile(fill="yellow", aes(alpha=gene.n==0)) + scale_alpha_manual(values=c(0,1), name="", labels=c('','gene desert')) + scale_fill_gradient(name="genes\nper Mb") + theme(legend.position=c(.99,.01), legend.justification=c(1,0)) + ggtitle("Protein-coding genes")
```

- Chr 19, 17 and 11 have more protein-coding genes than the rest.
- Chr Y has more pseudogene compared to other classes.

## Exons

### Number

Focusing on autosomes/X/Y, there are `r format(length(exons), big.mark=',')` "exons" from different types of genes:

```{r exontype}
exons %>% as.data.frame %>% mutate(gene_type=ifelse(grepl('IG|TR', gene_type), 'IG/TR', gene_type)) %>% group_by(gene_type) %>% summarize(vals=ceiling(n()/1000)) %>% dplyr::rename(names=gene_type) %>% arrange(desc(vals)) %>% as.data.frame %>% waffle(colors=pal[1:15], legend_pos='bottom', rows=20, xlab='1 sq = 1,000 exons')
exons = subset(exons, gene_type=="protein_coding")
```

For the rest of the analysis, I use only exons from protein-coding genes.

### Number per gene

```{r exongene}
exons.gene = exons %>% as.data.frame %>% group_by(gene_name) %>% summarize(exon=n())
exons.gene %>% mutate(exon.f=cut(exon, c(0,1,10,30,50,100,Inf))) %>% ggplot(aes(x=exon.f)) + geom_bar() + xlab('number of exons in gene') + ylab('number of genes') + theme_bw()
exons.gene %>% ungroup %>% summarize(mean.nb.exon=mean(exon), median.nb.exon=median(exon), max.nb.exon=max(exon)) %>% kable(digits=1)
```

The gene with the most exons is the [Titin gene](http://www.genecards.org/cgi-bin/carddisp.pl?gene=TTN):

```{r exongenetop}
exons.gene %>% arrange(desc(exon)) %>% head(5) %>% kable
```

### Size

The average exon size is `r round(mean(width(exons)))` bp. 

```{r exonsize}
exons %>% as.data.frame %>% ggplot(aes(x=width)) + geom_histogram() + ylab("number of exons") + xlab("exon length (bp)") + theme_bw() + scale_x_log10(breaks=c(1,10,100,1000,1e4))
genes10ex = exons %>% as.data.frame %>% filter(as.numeric(exon_number)>10) %>% .$gene_name %>% unique
exons %>% as.data.frame %>% filter(gene_name %in% genes10ex) %>% ggplot(aes(x=factor(winsor(as.numeric(exon_number), 10)), y=winsor(width, 1000))) + geom_violin() + geom_boxplot(alpha=0) + xlab("exon order") + ylab("exon length (bp)") + theme_bw() + scale_y_continuous(breaks=seq(0,1000,200), labels=c(seq(0,800,200), ">1000")) + scale_x_discrete(breaks=1:10, labels=c(1:9, "10+"))
```

The first exon seems to be slightly larger than the others. I used genes with at least 10 exons to be sure it's not due to large single-exon genes.

The largest annotated exons are more than 20 Kbp long:

```{r exonlarge}
exons %>% as.data.frame %>% arrange(desc(width)) %>% head(5) %>% mutate(size.kb=(width)/1e3) %>% select(seqnames, gene_name, size.kb) %>% kable(digits=2)
```

The smallest are just 1 bp !?

```{r exonsmall}
exons %>% as.data.frame %>% arrange(width) %>% head(5) %>% mutate(size.bp=width) %>% select(seqnames, gene_name, size.bp) %>% kable(digits=2)
```
