---
title: Alternative to pie charts in R
date: 2019-06-27
draft: true
tags: ["R", "plot"]
---

```{r, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, tidy=TRUE,
                      fig.width=10,
                      tidy.opts=list(width.cutoff=50))
```

## TL;DR

- **Pie charts ARE bad**: it's difficult to rank and compare the pies.
- **Waffle charts** are slightly better but it's still difficult to rank/compare groups.
- Good Ol' **bar charts are the best** way to quickly compare groups.
- We can make **more compact bar charts** with ggplot2 to use instead of pie charts.

## Set up

A typical tidyverse with data.frame manipulation (*dplyr*) and graphs (*ggplot2*).
Also, some functions to create random group labels and data.frames with labels that I will use to test the graphs.

```{r init}
library(ggplot2)
library(dplyr)

## Random names
randNames <- function(nb.names=10, name.length=6){
  replicate(nb.names, paste(sample(letters, name.length, TRUE), collapse=''))
}
## Creates a random data.frame with "class" column x
pieDF <- function(labels){
  tibble(x=rep(labels,
               c(runif(length(labels)/2, 100,200),
                 runif(length(labels)/2, 1,20))))
}
## Interleaves color for palette
interk <- function(x, k=4){ # Interleaves elements in x
  idx = unlist(lapply(1:k, function(kk) seq(kk, length(x), k)))
  x[idx]
}
pal = interk(rainbow(10, s=.8), 3)
```

I will **simulate 10 labels**: 5 with around the same number of elements (between 100 and 200); 5 with a small number of elements (between 1 and 20). 
This should be good to see how the graphs look like with multiple groups, including some with very low numbers.


## Problems with pie charts

Let's use **two datasets** representing counts of the same set of labels.
What can we say from a pie chart of one or both datasets?

```{r pie}
## Datasets
labs = randNames(10,6)
df1 = pieDF(labs)
df2 = pieDF(labs)
df12 = rbind(df1 %>% mutate(data='dataset 1'),
             df2 %>% mutate(data='dataset 2'))

ggplot(df1, aes(x='', fill=x)) + geom_bar(position='fill') +
  scale_fill_manual(values=pal) +
  coord_polar("y", start=0) + theme_bw() +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        legend.title=element_blank(),
        text=element_text(size=22))
```

Not that much. 
I see a couple of problems.

**Problem 1**: It's difficult to estimate the proportion of each group. 
Apparently humans are better at comparing straight lines than areas/pies (REF).

**Problem 2**: It's difficult to rank groups or to visualize differences between groups of similar sizes.

**Problem 3**: In this representation, we have to use the legend to match color with group label.
I don't like this when there are more than a couple groups.
To be fair, labels are often on top of the pies so it not an issue usually, just something to keep in mind below.

```{r pies}
ggplot(df12, aes(x='', fill=x)) + geom_bar(position='fill') +
  scale_fill_manual(values=pal) + 
  coord_polar("y", start=0) + theme_bw() +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        legend.title=element_blank(),
        text=element_text(size=22)) +
  facet_grid(.~data) 
```

**Problem 4**: It's a lot of back-and-forth work to understand how two datasets differ.


## Waffle graphs

Using [ggwaffle](https://github.com/liamgilbey/ggwaffle), let's make waffle graphs instead.

```{r waffle}
library(ggwaffle)

## Subset data to (better) control the number of waffle squares
waffleSubset <- function(df, ss=100){
  ss.prop = ss/nrow(df)
  df %>% group_by(x) %>%
    mutate(ss=round(n()*ss.prop), ss=ifelse(ss==0, 1, ss)) %>%
    sample_n(ss)
}

df1 %>% waffleSubset(200) %>% waffle_iron(aes_d(group=x), rows=10) %>%
  ggplot(aes(x, y, fill=group)) + geom_waffle() +
  coord_equal() + scale_fill_manual(values=pal) + theme_waffle() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.title=element_blank(),
        legend.position='bottom',
        text=element_text(size=22))
```

I find it a bit easier to visually rank and compare groups.
Still, problems 1, 2 and 3 are still present.

```{r waffles}
df12 %>% group_by(data) %>% do({waffleSubset(., 210) %>%
                                  waffle_iron(aes_d(group=x), rows=7)}) %>% 
  ggplot(aes(x, y, fill=group)) + geom_waffle() +
  facet_grid(data~.) + 
  coord_equal() + scale_fill_manual(values=pal) + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.title=element_blank(),
        panel.grid=element_blank(),
        panel.border=element_blank(),
        panel.background=element_blank(),
        axis.text=element_blank(),
        axis.line=element_blank(),
        plot.background=element_blank(),
        text=element_text(size=22))
```

Again, slightly better than pie charts (I think) but it's still not straightforward to compare the datasets.

## Bar graphs

A simple bar chart looks like this:

```{r bar}
ggplot(df1, aes(x=x, fill=x)) + geom_bar(color='black') +
  theme_bw() + scale_fill_manual(values=pal) +
  theme(axis.title.x=element_blank(), legend.title=element_blank()) +
  guides(fill=FALSE)
```

There is a lot of small text and white space in this graph.
It's fine as a graph, but we often want to use pie charts as a compact representation.

To make a **more compact bar chart**, we could:

- flip the graph
- reorder the bars
- increase the font size
- remove the legend and axis
- add the proportion as text annotation

```{r barc, fig.height=4, fig.width=5}
lab.digits=1
nudge=.005

df1 %>% group_by(x) %>% summarize(n=n()) %>% ungroup %>%
  mutate(prop=n/sum(n), percent=paste0(100*round(prop, 2+lab.digits),'%'),
         hjust=ifelse(prop<max(prop)/2, 0, 1),
         nudge=ifelse(prop<max(prop)/2, nudge, -1*nudge)) %>% 
  ggplot(aes(x=reorder(x, prop))) +
  geom_bar(aes(y=prop, fill=x), stat='identity', color='black') +
  geom_text(aes(y=prop+nudge, label=percent, hjust=hjust), size=7) + 
  theme_void() +
  scale_fill_manual(values=pal) +
  theme(axis.text.y=element_text(size=22)) + 
  coord_flip() + guides(fill=FALSE)
```

I like that much more: 

- the differences are easy to see visually.
- the numbers are there without taking space.
- the names of the groups are all in the same orientation and lined-up, making them easy to read.

```{r, barsc}
df12 %>% group_by(data, x) %>% summarize(n=n()) %>% ungroup %>%
    mutate(prop=n/sum(n), percent=paste0(100*round(prop, 2+lab.digits),'%'),
           hjust=ifelse(prop<max(prop)/2, 0, 1),
           nudge=ifelse(prop<max(prop)/2, nudge, -1*nudge)) %>% 
  ggplot(aes(x=reorder(x, prop, head, 1))) +
  geom_bar(aes(y=prop, fill=x), stat='identity', color='black') +
  geom_text(aes(y=prop+nudge, label=percent, hjust=hjust), size=7) + 
  theme_void() +
  scale_fill_manual(values=pal) +
  theme(axis.text.y=element_text(size=22), text=element_text(size=22)) + 
  coord_flip() + guides(fill=FALSE) +
  facet_grid(.~data)
```

When comparing two datasets, it is straightforward to align the groups.
In the end, I again think the differences are clearer to see.

Bar charts for the win!
