---
sidebar: true
weight: 1
title: R
output:
  blogdown::html_page:
    toc: true
---

## Details to remember

- Use `rbindlist` from *data.table* package for a memory-optimized and faster `do.call(rbind, list(..))`.
- Use `system2` instead of `system` to run a command. It's more portable apparently.
- Get RPubs working with `options(rpubs.upload.method = "internal")`.
- Operations on dates using `strptime(x, "%a %b %d %T PST %Y'")` and `difftime` and then `as.Double(d, units="hours")`.

### Order in condition assessment

Using `&` and `|` operators, *R* tries all the conditions and then performs the operations. 
However sometimes we would like a smarter sequential assessment for *AND*.
For example, we get an error if we run:

~~~r
x = NULL
if(!is.null(x) & x>10) message("so big !")
~~~

Because it tries to do `x>10` when *x* is *NULL*. 
Here, what we want is `&&`:

~~~r
x = NULL
if(!is.null(x) && x>10) message("so big !")
x = 17
if(!is.null(x) && x>10) message("so big !")
~~~

Now it won't try to do `x<10` if `!is.null(x)` is not true (because what's the point, anything "*False AND ...*" is for sure *False*).
Caution, `&&` doesn't work on vectors (it will only test the first element).

## Graphs

- `pdf("g.pdf", 9, 7)`
- `png("g.png", 1300, 1000, res=200)`
- `options(device=function() pdf(width=9, height=7))` to set the default device (e.g. remote graphs).

### Palette with many colors

When I need many colors and distinguish consecutive classes (in bar graphs or ordered clusters), I use an interleaved rainbow-like palette:

```{r, include=FALSE}
library(ggplot2)
bar.df = data.frame(x=sample.int(2, 100, TRUE), c=sample(letters[1:20],100, TRUE))
```

```{r}
interk <- function(x, k=4){ # Interleaves elements in x
  idx = unlist(lapply(1:k, function(kk) seq(kk,length(x), k)))
  x[idx]
}
pal = interk(rainbow(20, s=.8), 5)
```

For example:

```{r inter-palette, fig.height=2, fig.width=10}
ggplot(bar.df, aes(x=x, fill=c)) + geom_bar() + coord_flip() +
  guides(fill=guide_legend(ncol=4))
ggplot(bar.df, aes(x=x, fill=c)) + geom_bar() + coord_flip() +
  guides(fill=guide_legend(ncol=4)) + 
  scale_fill_manual(values=pal)
```

### ggplot2 tricks

- To plot a density distribution without the x-axis line, use `stat_density(geom="line")` (and eventually `position="dodge"` if plotting several groups).
- To override the legend's aes: `guides(colour=guide_legend(override.aes=list(alpha=1)))`
- Multi-column legends: `guides(fill=guide_legend(ncol=4))`

### Align/stack two ggplots

One easy way is to use the `tracks` function in the [ggbio](https://bioconductor.org/packages/release/bioc/html/ggbio.html) package. However I don't really like this package because it sometimes conflicts with ggplot2 (boo!) and you end up having to specify `ggplot2::` to the functions to avoid obscure errors.

I found [another way on the internet](http://www.exegetic.biz/blog/2015/05/r-recipe-aligning-axes-in-ggplot2/):

~~~r
library(ggplot2)
library(gridExtra)
p1 <- ggplot(...
p2 <- ggplot(...
p1 <- ggplot_gtable(ggplot_build(p1))
p2 <- ggplot_gtable(ggplot_build(p2))
maxWidth = unit.pmax(p1$widths[2:3], p2$widths[2:3])
p1$widths[2:3] <- maxWidth
p2$widths[2:3] <- maxWidth
grid.arrange(p1, p2, heights = c(3, 2))
~~~

### Change font in ggplot2

~~~r
library(extrafont)
font_import(pattern='Comic')
loadfonts()
qplot(x=rnorm(100)) + geom_histogram() + theme(text=element_text(family="Comic Sans MS")) + ggtitle('Ouch')
~~~

`fonts()` to check which fonts are imported by *extrafont*, `names(pdfFonts())` to list the fonts available (loaded).

More in [this blog post](http://blog.revolutionanalytics.com/2012/09/how-to-use-your-favorite-fonts-in-r-charts.html).

### Waffle graphs

`waffle` package provides a `waffle` and `iron` function. For example:

~~~r
iron(
  waffle(c(thing1=0, thing2=100), rows=5, keep=FALSE, size=0.5, colors=c("#af9139", "#544616")),
  waffle(c(thing1=25, thing2=75), rows=5, keep=FALSE, size=0.5, colors=c("#af9139", "#544616"))
)
~~~

## Rmarkdown

- `\ ` to force a line break and add vertical spacing (e.g. in slides).

To define *knitr* parameters, I add a chunk at the beginning of the Rmarkdown document. For example:

~~~md
```{r include=FALSE}`r ''`
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=10)
```
~~~

### Beamer presentation

Some useful options to put in the YAML header:

~~~yaml
title: The Title
subtitle: The Subtitle
author: Jean Monlong
date: 11 Oct. 2016
output:
  beamer_presentation:
    slide_level: 2
    fig_width: 7
    includes:
      in_header: header.tex
    toc: true
    dev: png
    keep_tex: true
~~~

+ `slide_level` defines the header level to be considered as a new slide.

To add slide count I put this on the `header.tex`:

~~~tex
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}[page number]
~~~

#### Wide tables

To resize wide tables I use a hook that surround a chunk with `\resizebox` command, defined in the non-included chunk:

~~~md
```{r, include=FALSE}`r ''`
knitr::knit_hooks$set(resize = function(before, options, envir) {
    if (before) {
      return('\\resizebox{\\textwidth}{!}{')
    } else {
      return('}')
    }
})
```

## Wide table

```{r, resize=TRUE}}`r ''`
knitr::kable(matrix(rnorm(10),10,10), format='latex')
```
~~~

### Jekyll website

The `Rmd` files located in the `_source` folder get automatically compiled by `servr` package using this command:

~~~sh
Rscript -e "servr::jekyll(script='build.R', serve=FALSE)"
~~~

Note: I now use [blogdown](https://bookdown.org/yihui/blogdown/) which automatically compile the R Markdown documents (every page is a R Markdown actually).

## *data.table* package

I'm more of a tidyverse person but for very large data the [data.table](https://cran.r-project.org/web/packages/data.table/) package is more efficient.

- [Cheatsheet from DataCamp](https://s3.amazonaws.com/assets.datacamp.com/blog_assets/datatable_Cheat_Sheet_R.pdf)

| *tidyverse* | *data.table* |
| --- | --- |
| `group_by(col1,col2) %>% summarize(nb=n())` | `dt[,.(nb=.N),by=.(col1,col2)]` |
| `group_by(col1,col2) %>% mutate(nb=n())` | `dt[,nb:=.N,by=.(col1,col2)]` |
| `filter(nb==2)` | `dt[nb==2]` |

## Linux setup

Necessary to compile some R packages:

~~~sh
sudo apt-get install libxml2-dev libssl-dev libmariadbclient-dev
~~~

*Related to XML, OpenSSL, MySQL, respectively.*
